---
title: "Chapter 8"
author: "Alan T. Arnholt"
date: 'Last compiled: `r format(Sys.time(), "%B %d, %Y at %X")`'
output: bookdown::html_document2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = NA, warning = FALSE, message = FALSE, fig.align = "center")
library(tidyverse)
library(janitor)
cereals <- read.csv("Cereals.csv") %>% clean_names()
db2014 <- read.csv("Dirt_bikes_2014.csv") %>% clean_names()
mod_cer <- lm(calories ~ sugars, data = cereals)
```

**Objectives:**

I.    Examine residuals

II.   Learn about leverage and influential observations

III.  Straightening scatterplots


# Examining Residuals

Read in the data set `Penguins.csv` using `read.csv()` and store the results in `penguins`.  Use the `clean_names()` function from the `janitor` package on `penguins` to get consistent variable names.  From this point forward, we will make a habit of using `clean_names()` for consistent variable names in our data frames.

```{r}
library(janitor)  # consistent variable names
penguins <- read.csv("Penguins.csv") %>% 
  clean_names()
knitr::kable(head(penguins))
```

Create a scatterplot of `dive_heart_rate` versus `duration_min` that resembles Figure 8.1 on page 240 of your text book.

```{r}
ggplot(data = penguins, aes(x = duration_min, y = dive_heart_rate)) + 
  geom_point(color = "brown") + 
  theme_bw() + 
  labs(x = "Duration (minutes)", y = "Dive heart rate (beats per minute)")
```
**Note** the shape of the relationship in the scatterplot.

Find the least squares line for regressing `dive_heart_rate` onto `duration_min` and store the results in `mod_pen`.

```{r}
mod_pen <- lm(dive_heart_rate ~ duration_min, data = penguins)
summary(mod_pen)
```

Plot the residuals for `mod_pen` versus the `duration_min` that resembles Figure 8.2 on page 240 of your text book.  Start by creating a new data frame (`NDF`) based on `mod_pen` using the `augment()` function from the `broom` package.

```{r}
library(broom)
NDF <- augment(mod_pen) %>% 
  clean_names()
knitr::kable(head(NDF))
```


```{r}
ggplot(data = NDF, aes(x = duration_min, y = resid)) +
  geom_point(color = "green4") + 
  theme_bw() + 
  geom_hline(yintercept = 0, lty = "dashed") + 
  labs(x = "Duration (min)", y = "Residuals")
```
Fix the model by transforming the response (take the `log` of `dive_heart_rate`) and create a scatterplot of the transformed response variable versus `duration_min`.

```{r}
# Your Code Goes HERE





```

Find the least squares estimates using the transformed response and store the results in `mod_pen2`

```{r}
penguins <- penguins %>% 
  mutate(log_dive_heart_rate = log(dive_heart_rate))
mod_pen2 <- lm(log_dive_heart_rate ~ duration_min, data = penguins)
summary(mod_pen2)
```

Plot the residuals of `mod_pen2` versus `duration_min`.

```{r}
# Your Code Goes HERE

```

Use `mod_pen2` to predict the heart rate of a penguin from a dive of 10 minutes.

```{r}
predict(mod_pen2, newdata = data.frame(duration_min = 10)) -> ans
ans
exp(ans)
```

Note: since `mod_pen2` is the `log` (which by default in R is the natural log (`ln`)), to get the answer back in the original units use `exp(ans)`.


## Sifting Residuals for Groups

Read in the data set `Cereals` and regress `calories` onto `sugars` and store the model in an object named `mod_cer`.

```{r}
# Your Code Goes HERE

```

Create a histogram of the residuals from `mod_cer` that resembles Figure 8.3 on page 241 of your text book.

```{r}
# Your Code Goes HERE

```

Using `mutate()` to create groups similar to Figure 8.4 on page 241 of your text book.

```{r}
NDF4 <- cereals %>%
  mutate(resid = residuals(mod_cer), fitted = fitted(mod_cer)) %>% 
  mutate(groups = ifelse(resid <= -25, "low", ifelse(resid <= 25, "med", "high")))
NDF4 %>% 
  filter(groups == "low") -> lowresid
NDF4 %>% 
  filter(groups == "high") -> highresid
knitr::kable(lowresid)
knitr::kable(highresid)
ggplot(data = NDF4, aes(x = fitted, y = resid, color = groups)) + 
  geom_point()+ 
  theme_bw()
```

Create a scatterplot of `calories` versus `sugars` color coded by `groups`.  Add the least squares lines for the three different color coded groups to the scatterplot.


```{r}
# Your Code Goes HERE
ggplot(data = NDF4, aes(x = sugars, y = calories, color = groups)) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE) + 
  theme_bw()
```

Explain the interaction and additive models below in class.

```{r}
mod_gro <- lm(calories ~ sugars*groups, data = NDF4)
summary(mod_gro)
# Note groups high is the base
mod_gro$coefficients
# LSE for high group
b0h <- mod_gro$coefficients[1]
b1h <- mod_gro$coefficients[2]
c(b0h, b1h)
# LSE for med group
b0m <- mod_gro$coefficients[1] + mod_gro$coefficients[4]
b1m <- mod_gro$coefficients[2] + mod_gro$coefficients[6]
c(b0m, b1m)
# LSE for low group
b0l <- mod_gro$coefficients[1] + mod_gro$coefficients[3]
b1l <- mod_gro$coefficients[2] + mod_gro$coefficients[5]
c(b0l, b1l)
ggplot(data = NDF4, aes(x = sugars, y = calories, color = groups)) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE) + 
  theme_bw() +
  geom_abline(slope = b1h, intercept = b0h, size = 0.25) + 
  geom_abline(slope = b1m, intercept = b0m, size = 0.25) + 
  geom_abline(slope = b1l, intercept = b0l, size = 0.25)
```

## Parallel slopes model

Load the `moderndive` package to use the `geom_parallel_slopes()` function to graph parallel lines.  Create a scatterplot of `calories` versus `sugars` color coded by `groups`.  Add parallel lines to the scatterplot.

```{r}
library(moderndive)
ggplot(data = NDF4, aes(x = sugars, y = calories, color = groups)) + 
  geom_point() + 
  geom_parallel_slopes(se = FALSE) + 
  theme_bw()
```
Explain the code below and how to get the intercepts and slopes for an additive model.

```{r}
# Parallel slope model
mod_ps <- lm(calories ~ sugars + groups, data = NDF4)
summary(mod_ps)
# Note groups high is the base
mod_ps$coefficients
# LSE for high group
b0h <- mod_ps$coefficients[1]
b1h <- mod_ps$coefficients[2]
c(b0h, b1h)
# LSE for med group
b0m <- mod_ps$coefficients[1] + mod_ps$coefficients[4]
b1m <- mod_ps$coefficients[2]
c(b0m, b1m)
# LSE for low group
b0l <- mod_ps$coefficients[1] + mod_ps$coefficients[3]
b1l <- mod_ps$coefficients[2]
c(b0l, b1l)
ggplot(data = NDF4, aes(x = sugars, y = calories, color = groups)) + 
  geom_point() + 
  geom_parallel_slopes(se = FALSE) + 
  theme_bw() +
  geom_abline(slope = b1h, intercept = b0h, size = 0.25) + 
  geom_abline(slope = b1m, intercept = b0m, size = 0.25) + 
  geom_abline(slope = b1l, intercept = b0l, size = 0.25)
```

## Book Figure 8.5

Recreate Figure 8.5 from page 242 of the text.

```{r}
cereals <- cereals %>% 
  mutate(shelf = ifelse(shelf == 1, "bottom", ifelse(shelf == 2, "middle", "top")))
ggplot(data = cereals, aes(x = sugars, y = calories, color = shelf)) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE) + 
  theme_bw() + 
  labs(x = "Sugar (g)", y = "Calories")
```

## Example 8.1 --- Extrapolation: Reaching Beyond the Data

Read in the `Marriage_age_2017.csv` data set and reproduce the analysis in Example 8.1.  Hint: use `mutate(time = ifelse(year <= 1940, "FP", ifelse(year <= 1979, "MP", "TP")))` to create a new variable `time` with three time periods.

```{r}
ma2017 <- read.csv("Marriage_age_2017.csv") %>% 
  clean_names()
ma2017 <- ma2017 %>% 
  mutate(time = ifelse(year <= 1940, "FP", ifelse(year <= 1979, "MP", "TP")))
# Your Code Goes HERE

```

## Outliers, Leverage, and Influence

Read in the `Election_2000.csv` data and recreate Figure 8.9 from page 246 of your text.

```{r}
election2000 <- read.csv("Election_2000.csv") %>% 
  clean_names()
p1 <- ggplot(data = election2000, aes(x = nader, y = buchanan, color = county)) + 
  geom_point() + 
  theme_bw() + 
  labs(x = "Nader (votes)", y = "Buchanan (votes)") + 
  guides(color=FALSE) # remove legend
p1
library(plotly) # create interactive graph
p2 <- ggplotly(p1)
p2
```

```{r}
election2000_PB <- election2000 %>% 
  filter(county != "PALM BEACH")
dim(election2000_PB)
mod_no_pb <- lm(buchanan ~ nader, data = election2000_PB)
dim(election2000)
ggplot(data = election2000, aes(x = nader, y = buchanan)) + 
  geom_point() + 
  theme_bw() + 
  labs(x = "Nader (votes)", y = "Buchanan (votes)") + 
  geom_smooth(method = "lm", se = FALSE) +
  geom_abline(intercept = coefficients(mod_no_pb)[1], slope = coefficients(mod_no_pb)[2])
```

## Lurking Variables and Causation

Read in the `Doctors_and_life_expectancy.csv` data set and recreate Figures 8.13 and 8.14 from page 249 of your text.

```{r}
dale <- read.csv("Doctors_and_life_expectancy.csv") %>% 
  clean_names()
names(dale)
ggplot(data = dale, aes( x = sqrt_doctors_person, y = life_exp)) + 
  geom_point(color = "brown") + 
  theme_bw() + 
  labs(x = expression(sqrt(Doctors/person)), y = "Life Expectancy (yr)", title = "Figure 8.13")
```

```{r}
ggplot(data = dale, aes( x = sqrt_tv_person, y = life_exp)) + 
  geom_point(color = "blue") + 
  theme_bw() + 
  labs(x = expression(sqrt(TVs/person)), y = "Life Expectancy (yr)", title = "Figure 8.14")
```

**Comment:** What should we do to increase life expectancy?


## Example 8.2

Read in the `Dirt_bikes_2014.csv` data set and store the result in `db2014`.  

```{r}
# Your Code Goes HERE

```

Create a scatterplot of `msrp` versus `displacement`.

```{r}
# Your Code Goes HERE

```

Regress `msrp` onto `displacement` and store the result in `mod_db`.  

```{r}
mod_db <- lm(msrp ~ displacement, data = db2014)
summary(mod_db)
```

Create a scatterplot of the residuals for `mod_db` versus the fitted values using the values in `NDF5`.

```{r}
NDF5 <- augment(mod_db) %>% 
  clean_names()
# Your Code Goes HERE

```

Add the variable `displacement_third` to the `db2014` data frame by raising `displacement` to the 1/3 power.

```{r}
db2014 <- db2014 %>% 
  mutate(displacement_third = displacement^(1/3))
```

Create a new data frame `NDF6` using the `filter()` function that contains only `Liquid` or `Air` cooled engines.

```{r}
NDF6 <- db2014 %>% 
  filter(air_cooled == "Liquid" | air_cooled == "Air")
```

Create a graph similar to the second graph of page 251 of your text using the data in `NDF6`.

```{r}
# Your Code Goes HERE

```

Modify your previous code from the previous graph to create a scatterplot with parallel lines for the different engine types.

```{r}
# Your Code Goes HERE

```

## Re-Expression for Regression

Read in the `Hand_dexterity.csv` data set and create a scatterplot of `time_sec` versus `age_yr` that resembles Figure 8.20 on page 254 of your text book.

```{r}
hd <- read.csv("Hand_dexterity.csv") %>% 
  clean_names() %>% 
  mutate(dominant = ifelse(dominant == 0, "Nondominant hand", "Dominant hand"))
# Your Code Goes HERE

```
Create a scatterplot of `speed` versus `age_yr` that resembles Figure 8.21 on page 254 of your text book.

```{r}
# Your Code Goes HERE

```

